Item Construction Rules Revisited: Learnings from Measurement of Latent Variables with Gold Standard Items

### 2025-08-05 manuscript.qmd
commit preprint version 1
commit preprint version 2 for PCI Psychology submission; pdf includes PCI Psychology checklist; keywords added

author-note:
  disclosures:
    data-sharing: This is version 0 (last modification {{< meta date-modified >}}) of our preprint published on PsyArXiv. All materials (reproducible manuscript, analysis code, datasets, codebooks, questionnaires) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/p7492/>. A Quarto Manuscript website is hosted at <https://florianpargent.github.io/gold-standard-items/>. The data has previously been analyzed in the dissertation thesis of the first author at <https://doi.org/10.5282/edoc.23628>. As part of the dissertation, a preregistration was published at <https://osf.io/cz3uv/?view_only=9cac02db231b48629fea9ae53c3038b9> that differs from the analyses reported in the current manuscript.
    conflict-of-interest: The authors declare that they have no financial conflicts of interest in relation to the content of the article, and have no non-financial conflicts of interest.
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
language:
  title-block-role-introduction: "Kathryn Eichhorn and Markus Bühner contributed equally to this manuscript. Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: "This article investigates whether gold-standard items are helpful for questionnaire construction. Based on an initial item pool (138 items) derived by a deductive item construction ($N$ = 124) and prototype approach ($N$ = 24), followed by cognitive interviews ($N$ = `r nrow(dat_interviews)`) and a pilot study ($N$ = `r nrow(dat_pilot)`) of a preliminary item set (61 items), three 12-item scales were constructed to measure the physical traits of body height, body weight, and age. We collected data on these scales using response formats with either two ($N$ = `r nrow(dat_dich)`) or six ($N$ = `r nrow(dat_poly)`) categories. We also collected numeric self-reports of body weight, body height, and age as gold-standard items. Confirmatory factor analyses revealed that the gold-standard items did not consistently exhibit the highest loading on their corresponding latent variable. Furthermore, when controlling for the self-reported physical body height, body weight, and age as gold-standard items, as well as gender, we did not always find an interpretable, systematic residual variance. Finally, the pattern of correlations between the latent variables did not reflect the correlations between the self-reported gold-standard items, suggesting that the item scales and the gold-standard items do not have the same validity. While these results are consistent with previous studies, our analyses also showed that items with two response categories were at least as valid as those with six categories, contradicting past recommendations. When constructing a questionnaire, we would argue that the items intended to measure the latent variable most directly should have the highest loading on that variable. If this is not the case, the content validity is questionable at best. The implications are that intensive cognitive pretesting is necessary. Questionnaires with different response formats should be compared empirically and hypotheses about which items best represent the latent variable should be tested."
license: CC BY 4.0
bibliography: [references.bib, r-references.bib]
floatsintext: true
numbered-lines: true
csl: apa.csl
---

>>>>

author-note:
  disclosures:
    data-sharing: This is version 2 (last modification {{< meta date-modified >}}) of our preprint published on PsyArXiv at <https://doi.org/10.31234/osf.io/r259d_v2>. All materials (reproducible manuscript, analysis code, datasets, codebooks, questionnaires) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/p7492/>. A Quarto Manuscript website is hosted at <https://florianpargent.github.io/gold-standard-items/>. The data has previously been analyzed in the dissertation thesis of the first author at <https://doi.org/10.5282/edoc.23628>. As part of the dissertation, a preregistration was published at <https://osf.io/cz3uv/?view_only=9cac02db231b48629fea9ae53c3038b9> that differs from the analyses reported in the current manuscript.
    conflict-of-interest: The authors declare that they have no financial conflicts of interest in relation to the content of the article, and have no non-financial conflicts of interest.
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
language:
  title-block-role-introduction: "Kathryn Eichhorn and Markus Bühner contributed equally to this manuscript. Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: "This article investigates whether gold-standard items are helpful for questionnaire construction. Based on an initial item pool (138 items) derived by a deductive item construction ($N$ = 124) and prototype approach ($N$ = 24), followed by cognitive interviews ($N$ = `r nrow(dat_interviews)`) and a pilot study ($N$ = `r nrow(dat_pilot)`) of a preliminary item set (61 items), three 12-item scales were constructed to measure the physical traits of body height, body weight, and age. We collected data on these scales using response formats with either two ($N$ = `r nrow(dat_dich)`) or six ($N$ = `r nrow(dat_poly)`) categories. We also collected numeric self-reports of body weight, body height, and age as gold-standard items. Confirmatory factor analyses revealed that the gold-standard items did not consistently exhibit the highest loading on their corresponding latent variable. Furthermore, when controlling for the self-reported physical body height, body weight, and age as gold-standard items, as well as gender, we did not always find an interpretable, systematic residual variance. Finally, the pattern of correlations between the latent variables did not reflect the correlations between the self-reported gold-standard items, suggesting that the item scales and the gold-standard items do not have the same validity. While these results are consistent with previous studies, our analyses also showed that items with two response categories were at least as valid as those with six categories, contradicting past recommendations. When constructing a questionnaire, we would argue that the items intended to measure the latent variable most directly should have the highest loading on that variable. If this is not the case, the content validity is questionable at best. The implications are that intensive cognitive pretesting is necessary. Questionnaires with different response formats should be compared empirically and hypotheses about which items best represent the latent variable should be tested."
keywords: [confirmatory factor analysis, content validity, item responses, latent variables, questionnaire development]
license: CC BY 4.0
bibliography: [references.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---
